---
title: "HNSCclassifier: An R Package to Predict Molecular Subtypes of Head and Neck Squamous Cell Carcinoma"
author: "Jiang Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
keywords: "HNSCC, classification, TCGA, ssGSEA, R package"
vignette: >
  %\VignetteIndexEntry{HNSCclassifier Introduction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Head and Neck Squamous Cell Carcinoma (HNSCC) is a molecularly
heterogeneous disease, and its classification into distinct subtypes —
**Atypical**, **Basal**, **Classical**, and **Mesenchymal** — has
important implications for prognosis and therapeutic strategy. The
`HNSCclassifier` package provides a ready-to-use tool for assigning
HNSCC samples to these TCGA-established molecular subtypes. The
classifier employs a **pathway-level normalization strategy** that first
converts gene expression profiles into pathway enrichment scores via
single-sample Gene Set Enrichment Analysis (ssGSEA), then applies a
pre-trained random forest model for final subtype prediction. This
design minimises cross-platform technical variation, making the
classifier applicable to both RNA-seq and microarray data.

## Installation

Before installing `HNSCclassifier`, ensure that the required
Bioconductor dependencies are available:

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("GSVA", "org.Hs.eg.db", "AnnotationDbi",
                       "randomForest", "shiny", "DT",
                       "shinythemes", "shinyjs"))
```

These packages provide the ssGSEA algorithm (`GSVA`) and the gene
annotation database (`org.Hs.eg.db`) used for identifier conversion. If
you plan to use the interactive Shiny interface, additionally install
`shiny`, `DT`, `shinythemes`, and `shinyjs`.

Once the dependencies are in place, install `HNSCclassifier` from
GitHub:

```{r, eval = FALSE}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("Ronlee12355/HNSCclassifier")
```

## Quick start

The classification workflow is straightforward: load the package,
prepare a numeric gene expression matrix with genes as rows and samples
as columns, and call `classifyHNSC()`. The function internally performs
input validation, optional log2 transformation, gene identifier
conversion (if needed), ssGSEA-based pathway scoring, and random forest
prediction — all in a single step. The `outputType` argument controls
whether the function returns predicted subtype **class labels** or a
**probability matrix** showing the posterior probability for each of the
four subtypes.

A built-in example dataset, `TCGA_LUSC`, is provided for demonstration.
It contains 50 TCGA lung squamous cell carcinoma samples with 22,962
genes in log2(TPM + 1) scale. While the dataset originates from a lung
cancer project, it serves as a convenient test case to verify that the
classifier runs correctly with your setup.

```{r, message = FALSE}
library(HNSCclassifier)

# Example: a toy expression matrix (genes as rows, samples as columns)
# Here we load a small built-in dataset for demonstration
data("TCGA_LUSC")

# Predict subtypes
subtypes <- classifyHNSC(TCGA_LUSC, outputType = "class")
table(subtypes)

# Get probabilities
probs <- classifyHNSC(TCGA_LUSC, outputType = "prob")
head(probs)
```

## Handling different gene ID types

Gene expression data from different bioinformatics pipelines often use
different gene identifier systems. The `classifyHNSC()` function
supports four common identifier types via the `idType` parameter:

- `"SYMBOL"` (default) — official HGNC gene symbols, e.g. *TP53*
- `"ENSEMBL"` — Ensembl gene IDs, e.g. *ENSG00000141510*
- `"ENTREZID"` — NCBI Entrez Gene IDs, e.g. *7157*
- `"REFSEQ"` — RefSeq accession numbers, e.g. *NM_000546*

When `idType` is not `"SYMBOL"`, the function automatically maps the
input identifiers to official gene symbols using the **org.Hs.eg.db**
annotation package. A common scenario where multiple input IDs map to
the same symbol is handled by retaining the **maximum expression value**
across the duplicates, thereby preserving the most biologically
informative signal. After conversion, the rest of the pipeline (log2
check, ssGSEA, random forest prediction) proceeds identically to
symbol-based input. The built-in dataset `TCGA_LUSC_ENSEMBL` provides a
ready-to-use example with Ensembl identifiers.

```{r, message = FALSE}
# For Ensembl IDs
res <- classifyHNSC(TCGA_LUSC_ENSEMBL, idType = "ENSEMBL")
table(res)
```

## Launching the Shiny interface

For users who prefer a graphical, point-and-click experience over
writing R code, the package includes an interactive Shiny web
application. The app provides a user-friendly interface for uploading
expression data, configuring classification parameters, and exploring
results — all within a web browser.

The app interface is organised into three tabs:

1. **Predict** — Upload a CSV or tab-delimited gene expression matrix
   (max 50 MB), select the gene identifier type and desired output
   format, and run the classifier. Results are displayed in an
   interactive table with copy, CSV, and Excel export options. A
   downloadable example file is provided to help users prepare their
   data in the correct format.
2. **Tutorial** — Step-by-step instructions on preparing an expression
   matrix, uploading data, and interpreting classification results.
3. **Contact** — Author information and a direct link to the GitHub
   Issues page for support.

The app performs the same underlying classification pipeline as the
command-line function, so results are identical regardless of the
interface chosen.

Ensure that **shiny**, **DT**, **shinythemes**, and **shinyjs** are
installed before launching:

```{r, eval = FALSE}
classifyHNSC_interface()
```

## Session info

```{r}
sessionInfo()
```
