---
title: "HNSCclassifier: An R Package to Predict Molecular Subtypes of Head and Neck Squamous Cell Carcinoma"
author: "Jiang Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
keywords: "HNSC, classification, TCGA, ssGSEA, R package"
vignette: >
  %\VignetteIndexEntry{HNSCclassifier Introduction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Head and Neck Squamous Cell Carcinoma (HNSC) is a molecularly
heterogeneous disease, and its classification into distinct subtypes —
**Atypical**, **Basal**, **Classical**, and **Mesenchymal** — has
important implications for prognosis and therapeutic strategy. The
`HNSCclassifier` package provides a ready-to-use tool for assigning
HNSC samples to these TCGA-established molecular subtypes. The
classifier employs a **pathway-level normalization strategy** that first
converts gene expression profiles into pathway enrichment scores via
single-sample Gene Set Enrichment Analysis (ssGSEA), then applies a
pre-trained random forest model for final subtype prediction. This
design minimises cross-platform technical variation, making the
classifier applicable to both RNA-seq and microarray data.

## Molecular Subtypes of HNSC

Head and Neck Squamous Cell Carcinoma is a heterogeneous disease with distinct molecular subtypes that differ in their biology, clinical outcomes, and therapeutic responses. The four subtypes, originally characterised by The Cancer Genome Atlas (TCGA) and Walter et al., are:

| Subtype | Key Characteristics |
|---|---|
| **Atypical** | Enriched for HPV-positive tumours; improved prognosis; p16<sup>INK4A</sup> overexpression; CDKN2A silencing less common |
| **Basal** | Expression patterns resembling basal epithelial cells; enrichments in epidermal development and extracellular matrix organisation genes |
| **Classical** | Heavy smoking association; the most prevalent subtype; characterised by xenobiotic metabolism, KEAP1/NRF2 pathway alterations, and oxidative stress gene signatures |
| **Mesenchymal** | Epithelial–mesenchymal transition (EMT) features; invasive/migratory phenotype; TGF-β signalling activation; poorest prognosis among the four subtypes |

Subtype information can guide prognosis stratification and may inform treatment selection in the context of clinical trials and translational research.

## Features

- **Command-line interface**: [`classifyHNSC()`](#quick-start) for batch prediction from a numeric gene expression matrix — ideal for processing large cohorts or integrating into downstream pipelines.
- **Interactive Shiny app**: [`classifyHNSC_interface()`](#shiny-web-interface) for users who prefer a graphical, point-and-click experience.
- **Multi-platform identifier support**: accepts gene symbols, Ensembl IDs, Entrez IDs, or RefSeq accession numbers. Automatic conversion is performed via `org.Hs.eg.db`.
- **Flexible output**: return predicted subtype **class labels** (`"class"`) or a **probability matrix** (`"prob"`) showing posterior probabilities for each of the four subtypes.
- **Cross-platform robustness**: designed to work with RNA-seq (TPM, FPKM, or normalised counts) and microarray expression data after log2 transformation.
- **Automatic preprocessing**: input validation, log2 transformation detection and application, and gene-level noise reduction (zero-variance gene filtering) are handled internally.
- **Curated pathway gene sets**: built with a carefully selected collection of pathway gene sets (`required.sets`) for ssGSEA, bridging any input dataset to the model's training feature space.
- **Pre-trained random forest model**: ships with a fully trained `randomForest` object (`finalModel`) — no need for users to train or tune any model.


## Installation

Before installing `HNSCclassifier`, ensure that the required
Bioconductor dependencies are available:

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("GSVA", "org.Hs.eg.db", "AnnotationDbi",
                       "randomForest", "shiny", "DT",
                       "shinythemes", "shinyjs"))
```

These packages provide the ssGSEA algorithm (`GSVA`) and the gene
annotation database (`org.Hs.eg.db`) used for identifier conversion. If
you plan to use the interactive Shiny interface, additionally install
`shiny`, `DT`, `shinythemes`, and `shinyjs`.

Once the dependencies are in place, install `HNSCclassifier` from
GitHub:

```{r, eval = FALSE}
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("Ronlee12355/HNSCclassifier")
```

## Quick start

The classification workflow is straightforward: load the package,
prepare a numeric gene expression matrix with genes as rows and samples
as columns, and call `classifyHNSC()`. The function internally performs
input validation, optional log2 transformation, gene identifier
conversion (if needed), ssGSEA-based pathway scoring, and random forest
prediction — all in a single step. The `outputType` argument controls
whether the function returns predicted subtype **class labels** or a
**probability matrix** showing the posterior probability for each of the
four subtypes.

A built-in example dataset, `TCGA_LUSC`, is provided for demonstration.
It contains 50 TCGA lung squamous cell carcinoma samples with 22,962
genes in log2(TPM + 1) scale. While the dataset originates from a lung
cancer project, it serves as a convenient test case to verify that the
classifier runs correctly with your setup.

```{r, message = FALSE}
library(HNSCclassifier)

# Example: a toy expression matrix (genes as rows, samples as columns)
# Here we load a small built-in dataset for demonstration
data("TCGA_LUSC")

# Predict subtypes
subtypes <- classifyHNSC(TCGA_LUSC, outputType = "class")
table(subtypes)

# Get probabilities
probs <- classifyHNSC(TCGA_LUSC, outputType = "prob")
head(probs)
```

## Handling different gene ID types

Gene expression data from different bioinformatics pipelines often use
different gene identifier systems. The `classifyHNSC()` function
supports four common identifier types via the `idType` parameter:

- `"SYMBOL"` (default) — official HGNC gene symbols, e.g. *TP53*
- `"ENSEMBL"` — Ensembl gene IDs, e.g. *ENSG00000141510*
- `"ENTREZID"` — NCBI Entrez Gene IDs, e.g. *7157*
- `"REFSEQ"` — RefSeq accession numbers, e.g. *NM_000546*

When `idType` is not `"SYMBOL"`, the function automatically maps the
input identifiers to official gene symbols using the **org.Hs.eg.db**
annotation package. A common scenario where multiple input IDs map to
the same symbol is handled by retaining the **maximum expression value**
across the duplicates, thereby preserving the most biologically
informative signal. After conversion, the rest of the pipeline (log2
check, ssGSEA, random forest prediction) proceeds identically to
symbol-based input. The built-in dataset `TCGA_LUSC_ENSEMBL` provides a
ready-to-use example with Ensembl identifiers.

```{r, message = FALSE}
# For Ensembl IDs
res <- classifyHNSC(TCGA_LUSC_ENSEMBL, idType = "ENSEMBL")
table(res)
```

## Extracting pathway-level enrichment scores

For advanced users who wish to perform their own downstream analyses
(e.g. pathway-level differential expression, clustering, or custom
machine learning), the `pathway_scores()` function exposes the internal
ssGSEA enrichment scores that the random forest classifier uses as
features. It runs the same preprocessing pipeline as `classifyHNSC()`
— input validation, gene ID conversion, log2 transformation, and ssGSEA
scoring — but returns the pathway × sample enrichment matrix instead of
predicted subtypes.

```{r, message = FALSE}
pw <- pathway_scores(TCGA_LUSC)
dim(pw)
head(pw[, 1:5])
```

The returned matrix can be used with standard bioinformatics tools such
as `limma` for differential pathway analysis, `ConsensusClusterPlus` for
unsupervised clustering, or `pheatmap` for visualisation.

## Visualising posterior subtype probabilities

The `plot_subtype_probabilities()` function creates a stacked bar chart
showing the posterior probability of each subtype for every sample.
Samples are grouped by predicted subtype and ordered by confidence
within each group. Samples whose maximum probability falls below a
user-defined threshold (default 0.5) are flagged with a red asterisk,
helping identify borderline cases that may warrant closer inspection.

```{r, fig.width = 6, fig.height = 4, out.width = "100%", message = FALSE}
probs <- classifyHNSC(TCGA_LUSC, outputType = "prob")
plot_subtype_probabilities(probs)
```

To adjust the low-confidence threshold:

```{r, fig.width = 6, fig.height = 4, out.width = "100%", message = FALSE}
plot_subtype_probabilities(probs, threshold = 0.7)
```

## Visualising subtype-specific gene expression

The `plot_subtype_heatmap()` function generates a heatmap of curated
subtype-driver genes, with samples annotated by their predicted
molecular subtype. It uses the built-in `subtype_genes` set, and the
expression values are row-wise z-score scaled for comparability.

```{r, fig.width = 8, fig.height = 8, out.width = "100%", message = FALSE}
subtypes <- classifyHNSC(TCGA_LUSC, outputType = "class")
plot_subtype_heatmap(TCGA_LUSC, subtypes)
```

A custom gene set can be supplied via the `gene_set` argument, and
clustering behaviour can be controlled with `cluster_rows` and
`cluster_cols`.

## Working with full TCGA-LUSC data

The built-in `TCGA_LUSC` dataset is a convenience subset of 50 samples
for quick testing.  For real analyses, you can download the full
TCGA-LUSC expression cohort and reproduce the complete classification
workflow.  This section walks through each step — from downloading raw
data to comparing predictions against the original TCGA consensus
annotations.

### Downloading expression data

The TCGA-LUSC gene expression matrix (STAR-aligned, log2(TPM+1)
transformed) is hosted on the [UCSC Xena data
browser](https://xenabrowser.net/datapages/).  Download the file
`TCGA-LUSC.star_tpm.tsv.gz`:

```{bash, eval = FALSE}
wget https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-LUSC.star_tpm.tsv.gz
gunzip TCGA-LUSC.star_tpm.tsv.gz
```

### Downloading TCGA subtype annotations

The TCGA molecular subtype calls are available via
**TCGAbiolinks**.  Install it if needed:

```{r, eval = FALSE}
BiocManager::install("TCGAbiolinks")
```

### Full pipeline: from raw data to predicted subtypes

The following code reads the expression TSV, strips Ensembl version
suffixes, converts Ensembl IDs to gene symbols, retains primary tumour
samples only, filters zero-variance genes, and runs `classifyHNSC()`.
It also fetches the TCGA reference annotation for downstream comparison.

```{r, eval = FALSE}
library(data.table)
library(org.Hs.eg.db)
library(dplyr)
library(tibble)
library(stringr)
library(TCGAbiolinks)
library(HNSCclassifier)

# ---- 1. Read expression data ----
lusc.expr <- data.table::fread(
  "TCGA-LUSC.star_tpm.tsv",
  data.table = FALSE
)

# Strip Ensembl version suffix (e.g. ENSG00000141510.11 → ENSG00000141510)
lusc.expr$Ensembl_ID <- sapply(
  lusc.expr$Ensembl_ID,
  function(x) unlist(strsplit(x, "\\."))[1]
)

# Convert Ensembl ID → gene symbol
lusc.expr$symbol <- AnnotationDbi::mapIds(
  org.Hs.eg.db,
  keys      = lusc.expr$Ensembl_ID,
  keytype   = "ENSEMBL",
  column    = "SYMBOL"
)

lusc.expr$Ensembl_ID <- NULL
lusc.expr <- lusc.expr %>%
  dplyr::filter(!is.na(symbol)) %>%
  dplyr::group_by(symbol) %>%
  dplyr::summarise_all(max) %>%
  dplyr::ungroup() %>%
  as.data.frame() %>%
  tibble::column_to_rownames("symbol")

# Keep primary tumour samples only (barcodes ending with "-01A")
lusc.expr <- lusc.expr[, grepl("01A$", colnames(lusc.expr))]

# ---- 2. TCGA subtype annotation ----
lusc.subtype <- TCGAbiolinks::TCGAquery_subtype(tumor = "LUSC")
lusc.subtype$sample <- paste0(lusc.subtype$patient, "-01A")
lusc.subtype <- lusc.subtype[
  lusc.subtype$sample %in% colnames(lusc.expr),
]
lusc.expr <- lusc.expr[, lusc.subtype$sample]

# Harmonise subtype labels to title case
lusc.subtype$Expression.Subtype <- stringr::str_to_title(
  lusc.subtype$Expression.Subtype
)

# Remove zero-variance genes
lusc.expr <- lusc.expr[apply(lusc.expr, 1, mad) > 0, ]

# ---- 3. Run HNSCclassifier ----
lusc_pred <- classifyHNSC(lusc.expr, outputType = "class")
table(lusc_pred)
```

### Comparing with TCGA consensus annotations

The TCGA project assigned HNSC subtypes by integrating mRNA, miRNA,
methylation, and copy-number data via consensus clustering.  The
`HNSCclassifier` uses only transcriptomic data, so perfect agreement
is not expected; nevertheless, a cross-tabulation gives a useful sense
of concordance:

```{r, eval = FALSE}
comp <- data.frame(
  TCGA   = lusc.subtype$Expression.Subtype,
  HNSCclass = lusc_pred[lusc.subtype$sample]
)
table(comp)
```

## Launching the Shiny interface

For users who prefer a graphical, point-and-click experience over
writing R code, the package includes an interactive Shiny web
application. The app provides a user-friendly interface for uploading
expression data, configuring classification parameters, and exploring
results — all within a web browser.

The app interface is organised into three tabs:

1. **Predict** — Upload a CSV or tab-delimited gene expression matrix
   (max 50 MB), select the gene identifier type and desired output
   format, and run the classifier. Results are displayed in an
   interactive table with copy, CSV, and Excel export options. A
   downloadable example file is provided to help users prepare their
   data in the correct format.
2. **Tutorial** — Step-by-step instructions on preparing an expression
   matrix, uploading data, and interpreting classification results.
3. **Contact** — Author information and a direct link to the GitHub
   Issues page for support.

The app performs the same underlying classification pipeline as the
command-line function, so results are identical regardless of the
interface chosen.

Ensure that **shiny**, **DT**, **shinythemes**, and **shinyjs** are
installed before launching:

```{r, eval = FALSE}
classifyHNSC_interface()
```

## Session info

```{r}
sessionInfo()
```
