---
title: "HNSCclassifier: An R Package to Predict Molecular Subtypes of Head and Neck Squamous Cell Carcinoma"
author: "Jiang Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
keywords: "HNSCC, classification, TCGA, ssGSEA, R package"
vignette: >
  %\VignetteIndexEntry{HNSCclassifier Introduction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `HNSCclassifier` package provides a ready-to-use tool for
classifying head and neck squamous cell carcinoma samples into
the established TCGA molecular subtypes (Atypical, Basal, Classical,
Mesenchymal). It uses a pathway-level normalization strategy based on
ssGSEA and a pre-trained random forest model.

## Installation

```{r, eval = FALSE}
# Install from GitHub
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("Ronlee12355/HNSCclassifier")
```

## Quick start

To classify your HNSCC samples, load the package, prepare a gene
expression matrix with gene symbols in rows and sample identifiers in
columns, and call `classifyHNSC()`. The function returns either the
predicted molecular subtype for each sample or a matrix of class
probabilities, depending on the `outputType` argument. A small built-in
example dataset is included to illustrate the typical workflow.

```{r, message = FALSE}
library(HNSCclassifier)

# Example: a toy expression matrix (genes as rows, samples as columns)
# Here we load a small built-in dataset for demonstration
data("TCGA_LUSC")

# Predict subtypes
subtypes <- classifyHNSC(TCGA_LUSC, outputType = "class")
table(subtypes)

# Get probabilities
probs <- classifyHNSC(TCGA_LUSC, outputType = "prob")
head(probs)
```

## Handling different gene ID types

If the rownames of your expression matrix are not gene symbols, you can
use the `idType` argument to specify the identifier type (e.g.
`"ENSEMBL"`, `"ENTREZID"`, or `"REFSEQ"`). The function will
automatically convert them to official gene symbols using the
**org.Hs.eg.db** annotation package (which must be installed
beforehand). When multiple IDs map to the same gene symbol, the maximum
expression value is retained. After conversion, the classification
proceeds exactly as with symbol-based input.

```{r, message = FALSE}
# For Ensembl IDs
res <- classifyHNSC(TCGA_LUSC_ENSEMBL, idType = "ENSEMBL")
table(res)
```

## Launching the Shiny interface

For users who prefer a point-and-click interface, the package includes
an interactive Shiny web application. It accepts gene expression
matrices in CSV or tab-delimited format, automatically handles ID
conversion and classification, and displays both subtype assignments and
probability heatmaps. Ensure that **shiny** and **DT** are installed,
then launch the app with:

```{r, eval = FALSE}
classifyHNSC_interface()
```

## Session info

```{r}
sessionInfo()
```
